A Recipe for Scavenging \textcolor{blue}{in Vertebrates} - the natural history of a behaviour

### 2016-07-20 natural history of scavenging.tex
Kevin updates
Kevin updates

Historically, \textcolor{blue}{food webs have classified most acts of carnivory as predation events \citep{wilson2011scavenging}. 
This is clearly a mistaken view because of the existence of scavenging, a behaviour displayed by almost all carnivorous vertebrates \citep{devault2003scavenging}. 
Recent research has begun to redress this imbalance by recognising the prevalence of scavenging and detailing its ecological consequences across modern ecosystems such that are beginning to appreciate the implications for trophic ecology \citep{pereira2014facultative,periquet2015lion}.
Indeed, there is no discrete divide between predators and scavengers but rather a continuous gradient in terms of the proportion of carrion in the diet \citep{pereira2014facultative}.
Even vultures, the canonical example of obligate scavengers, can hunt \citep{margalida2011scavenger}. 
Unfortunately, the lack of direct behavioural data makes it difficult to discern scavenging from predation among extinct forms where it was surely exhibited by prehistoric vertebrate carnivores.
Some simple heuristics can inform us whether or not the prey was scavenged, for instance, in cases where it was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}.}
But clearly, a scavenger does not only feed on animals too big for it to have hunted.

@@

\section{Encounter Rate}
All foraging processes depend on the encounter rate between consumer and resource. 
Locomotory speed, foraging time and detection radius all determine the encounter rate between a scavenger and the carcasses it is searching for and we would thus expect selection pressures to act on the various traits that govern these parameters.
Selection might also drive a species to reduce its metabolic requirements so that it can survive long periods between meals, or expand a species' ability to process tissues or decaying flesh that might ordinarily be discarded.
However, as we noted above, encounter rate is also determined by the productivity of the environment, and ultimately the rate at which carcasses are produced. 
\textcolor{blue}{Scavengers can drive this process indirectly, for instance, by consuming the majority of carrion, more obligate scavengers deny a resource to their facultative counterparts obliging the latter to hunt more, and thus, increase the amount of carrion \citep{moleon2014inter}.}

@@

@@

Of course, terrestrial animals can also move bipedally. 
Although the evolution of bipedal movement was significant in that it freed up the forelimbs for other purposes (e.g. climbing, tool-use, wing development etc.) it does not differ radically in cost from quadrupedal locomotion (\cite{williams1999evolution}, and references therein). 
For instance, \cite{alexander2004bipedal} shows that, in the case of humans, we are more economical than predicted while walking and less so while running according to predicted costs of terrestrial movement calculated by allometric scaling relationships.
Our locomotory efficiency has fed into the question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene, which has been a matter of debate for years \citep{dominguez2002hunting}.
\cite{ruxton2013endurance} added to this debate with their argument that long distance endurance running \textcolor{blue}{was not an important feature of hominin scavenging but was instead used by humans for hunting.
Even when considering the use of weapons and social foraging, the authors argue that the high costs of running long distances would render it an unattractive foraging strategy for scavenging.}

Aside from humans and our allies, the best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been extensively explored in the carnivorous theropods.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
While the locomotory ability of theropods has been debated since their first inception, more recent studies have reconstructed them as relatively mobile animals \citep{pontzer2009biomechanics}. 
Despite some suggestions that larger species may have had some advantage in scavenging, partially due to the ability to search large areas \citep{ruxton2003could}, more recent work has shown that the energetic demands of such large forms meant scavenging was likely more prevalent in mid-sized theropods of approximately half a tonne \citep{kane2016body}.

@@

\textcolor{blue}{The physical differences between water and air mean carcass availability is radically different between these environments \citep{beasley2012carrion}. 
For one, carcasses get moved around by water which results in a more diffuse signal being produced for would-be scavengers.
Carcasses also tend to sink in water where they are no longer accessible to pelagic scavengers \citep{beasley2012carrion}. 
\textcolor{blue}{Research has shown that an animal need only travel 36 km to encounter a fresh whale carcass \citep{smith2003ecology}.} 
The phenomenon of occasional bounties of carrion in the form of these whale falls has led some researchers to investigate if an obligate scavenger could survive by seeking out these remains exclusively.}
\cite{ruxton2005searching} argued that although this is energetically feasible it's ecologically unlikely.
Any animal that could find such whale carcasses is unlikely to have ignored other types of carrion.
Although no aquatic species have ever exceeded the size of whales, some enormous animals have evolved in this environment before the evolution of cetaceans, including \textit{Leedsichthys}, a bony fish from the Middle Jurassic (174.1 - 163.5 Mya) and the aquatic Mesozoic reptiles, the plesiosaurs, pliosaurs and ichtyosaurs, that could all exceed 15 metres in length \citep{ruxton2011zoology,danise2014ecological}.
So, despite being unlikely, the energetic feasibility of a marine scavenger that specialises on large carcasses has a long history.

Perhaps the greatest environmental driver of scavenging tendency is that of temperature. 
This has a significant bearing on the availability and persistence of carrion.
We know from the geological record shows that the Earth has undergone radical fluctuations in temperature over time.
\textcolor{blue}{On land, there are a wide variety of vegetation types, from thickly forested areas to open grassland and the extent of this vegetation changes with season.}
All of these factors influence the availabiltiy of carrion.  
Moreover, oceanic productivity is also impacted by climactic conditions.

To illustrate the point, a 10$^{\circ}$C increase in ambient temperature can double carcass decomposition rates \citep{parmenter2009carrion} and geological evidence indicates that the Mesozoic Earth was on average at least 6 $^{\circ}$C warmer than now \citep{sellwood2006mesozoic}.
In terms of specific habitats, it has been shown that decomposition is greater in warm and moist areas versus more xeric ones \citep{beasley2015vertebrates}.
The impacts these can have on scavengers have been empirically supported e.g. \cite{beasley2015vertebrates} who point to a series of studies showing how microbes and invertebrates benefit at higher temperatures to the detriment of vertebrate scavengers such that ``above 20$^{\circ}$C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18$^{\circ}$C, vertebrates consumed 49\% of carcasses".
\textcolor{blue}{In savannah, dense vegetation can prevent vultures from landing to a carcass because they will be unable to take off again \citep{bamford2009effect}. 
The Earth has also undergone a series of ice ages of various spatio-temporal extent \citep{diedrich2012cave} and modern terrestrial settings that experience sub-zero conditions can act as a microcosm to show the effect extreme cold can have on scavenging.
Notably, freezing carcasses can become too hard to consume by most vertebrate carnviores \citep{selva2003scavenging}.}

@@

In addition to fending off other vertebrates, scavengers also have to contend with competition from micro-organisms, which requires a specialised physiology. 
Although the findings of \cite{shivik2006vultures} that ``evolutionary pressures favor detection maximizers relative to toxification minimizers in competitive interactions for carcasses." appear sound, the fact remains that overcoming micro-organism toxins is still a beneficial adaptation to any scavenger. 
Avian scavengers have evolved incredibly acidic stomachs that allow them to consume and process putrefied flesh with no ill effects \citep{houston1975digestive,roggenbuck2014microbiome}. 
This adaptation is not restricted to vultures though, \cite{gremillet2012vultures} showed wandering albatrosses (\textit{Diomedea exulans}; so-called ``vultures of the seas'') had an average pH of 1.5, which enables them to consume fisheries discards and squid carcasses. 
There is also evidence of selection for ``toxification minimizers'' beyond birds among the ectotherms.
From our earlier arguments we know that ecthotherms are limited in their ability to find carrion as quickly as endotherms. 
These later arrivers would thus benefit especially from well-developed detoxifying apparatus. 
\cite{shivik2006vultures} suggests that ``specialized oral structures in snakes may have evolved under pressures associated with scavenging."
Moreover, some researchers have charted an evolutionary course from basal fossorial snakes to modern terrestrial species by way of an obligate scavenger intermediate \citep{bauchot2006snakes}. 
\textcolor{blue}{In water, by contrast, competition with micro-organisms is significantly reduced because of the high pressures and low temperatures \citep{beasley2012carrion}.}

>>>>

Historically, \textcolor{blue}{food webs have classified most acts of carnivory as predation events \citep{wilson2011scavenging}. 
This is clearly a mistaken view because of the existence of scavenging, a behaviour displayed by almost all carnivorous vertebrates \citep{devault2003scavenging}. 
Recent research has begun to redress this imbalance by recognising the prevalence of scavenging and its implications for trophic ecology through detailing the ecological consequences of scavanging across modern ecosystems \citep{pereira2014facultative,periquet2015lion}.
Indeed, there is no discrete divide between predators and scavengers but rather a continuous gradient in terms of the proportion of carrion in the diet \citep{pereira2014facultative}.
Even vultures, the canonical example of obligate scavengers, can hunt \citep{margalida2011scavenger}. While it is relatively easy to determine whether a species engages in scavenging through observation, determining proportions requires extensive behavioral data (I put this in to placate the reviewer who said it was easy).
Unfortunately, the lack of direct behavioural data makes it even more difficult to discern scavenging from predation among extinct forms where it was surely exhibited by prehistoric vertebrate carnivores.
Some simple heuristics can inform us whether or not the prey was scavenged, for instance, in cases where it was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}.}
But clearly, a scavenger does not only feed on animals too big for it to have hunted.

@@

\section{Encounter Rate}
All foraging processes depend on the encounter rate between consumer and resource. 
Locomotory speed, foraging time and detection radius all determine the encounter rate between a scavenger and the carcasses it is searching for and we would thus expect selection pressures to act on the various traits that govern these parameters.
Selection might also drive a species to reduce its metabolic requirements so that it can survive long periods between meals, or expand a species' ability to process tissues or decaying flesh that might ordinarily be discarded.
However, as we noted above, encounter rate is also determined by the productivity of the environment, and ultimately the rate at which carcasses are produced. 
\textcolor{blue}{Carcass production itself depends on factors such as predation, disease and scavenging itself. For instance, by consuming the majority of carrion, more obligate scavengers deny a resource to their facultative counterparts obliging the latter to hunt more, and thus, increase the amount of carrion \citep{moleon2014inter}.}

@@

\textcolor{blue}{While it is difficult to infer the metabolisms of extinct species, similarities can still also be drawn. Mainly ectothermic groups such as sharks and bony fish were likely to benefit from the ability to bridge long time-spans between meals such as seen in snakes. In contrast, forms such as dinosaurs \citep{grady2014evidence}, pterosuars \citep{wellnhofer1991illustrated}, some marine reptiles \citep{bernard2010regulation} and some early mammals \citep{lovegrove2016phenology} have been proposed to have at least partial endothermy resulting in requiring a higher intake of material in comparison to their ectothermic equivalents. However, such higher metabolic rates would also increased the ability of such species to actively search for carrion.} \textcolor{red}{I put this section in as there is no reference in how we can use the above paragraph to infer scavenging in extinct species}

@@

Of course, terrestrial animals can also move bipedally. 
Although the evolution of bipedal movement was significant in that it freed up the forelimbs for other purposes (e.g. climbing, tool-use, wing development etc.) it does not differ radically in cost from quadrupedal locomotion (\cite{williams1999evolution}, \textcolor{blue}{and references therein}). 
For instance, \cite{alexander2004bipedal} shows that, in the case of humans, we are more economical than predicted while walking and less so while running according to predicted costs of terrestrial movement calculated by allometric scaling relationships.
Our locomotory efficiency has fed into the question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene, which has been a matter of debate for years \citep{dominguez2002hunting}.
\cite{ruxton2013endurance} added to this debate with their argument that long distance endurance running \textcolor{blue}{was not an important feature of hominin scavenging but was instead used by humans for hunting.
Even when considering the use of weapons and social foraging, the authors argue that the high costs of running long distances would render it an unattractive foraging strategy for scavenging.}

Aside from humans and our allies, the best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been extensively explored in the carnivorous theropods.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
While the locomotory ability of theropods has been debated since their first inception, more recent studies have reconstructed them as relatively mobile animals \citep{pontzer2009biomechanics}. 
Despite some suggestions that larger species may have had some advantage in scavenging, partially due to the ability to search large areas \citep{ruxton2003could}, more recent work has shown that the energetic demands of locomotion in such large forms meant scavenging was likely more prevalent in mid-sized theropods of approximately half a tonne \citep{kane2016body}.

@@

\textcolor{blue}{The physical differences between water and air mean carcass availability is radically different between these environments \citep{beasley2012carrion}. 
For one, carcasses get moved around by water which results in a more diffuse signal being produced for would-be scavengers.
Carcasses also tend to sink in water where they are no longer accessible to pelagic scavengers \citep{beasley2012carrion}. 
\textcolor{blue}{Research has shown that an animal need only travel 36 km to encounter a fresh whale carcass \citep{smith2003ecology}.} 
The phenomenon of occasional bounties of carrion in the form of these whale falls has led some researchers to investigate if an obligate scavenger could survive by seeking out these remains exclusively.}
\cite{ruxton2005searching} argued that although this is energetically feasible it's ecologically unlikely.
Any animal that could find such whale carcasses is unlikely to have ignored other types of carrion.
Although no aquatic species have ever exceeded the size of whales, some enormous animals have evolved in this environment before the evolution of cetaceans, including \textit{Leedsichthys}, a bony fish from the Middle Jurassic (174.1 - 163.5 Mya) and the aquatic Mesozoic reptiles, the plesiosaurs, pliosaurs and ichtyosaurs, that could all exceed 15 metres in length \citep{ruxton2011zoology,danise2014ecological}.
So, despite being unlikely \textcolor{red}{Is there an estimate of how much of the fishes diet can be made up from this type of scavenging?}, the energetic feasibility of a marine scavenger that specialises on large carcasses has a long history.

Perhaps the greatest environmental driver of scavenging tendency is that of temperature. 
This has a significant bearing on the availability and persistence of carrion.
We know from the geological record that the Earth has undergone radical fluctuations in temperature over time.
\textcolor{blue}{On land, there are a wide variety of vegetation types, from thickly forested areas to open grassland and the extent of this vegetation changes with season.}
All of these factors influence the availabiltiy of carrion.  
Moreover, oceanic productivity is also impacted by climactic conditions. \textcolor{red}{I would delete the last two lines of this paragraph and merge it with the next one}

To illustrate the point, a 10$^{\circ}$C increase in ambient temperature can double carcass decomposition rates \citep{parmenter2009carrion} and geological evidence indicates that the Mesozoic Earth was on average at least 6 $^{\circ}$C warmer than now \citep{sellwood2006mesozoic}.
In terms of specific habitats, it has been shown that decomposition is greater in warm and moist areas versus more xeric ones \citep{beasley2015vertebrates}.
The impacts these can have on scavengers have been empirically supported e.g. \cite{beasley2015vertebrates} who point to a series of studies showing how microbes and invertebrates benefit at higher temperatures to the detriment of vertebrate scavengers such that ``above 20$^{\circ}$C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18$^{\circ}$C, vertebrates consumed 49\% of carcasses".
\textcolor{blue}{In savannah, dense vegetation can prevent vultures from landing to a carcass because they will be unable to take off again \citep{bamford2009effect} \textcolor{red}{This is a biotic interaction so I would either remove it or move it somewhere else}. 
The Earth has also undergone a series of ice ages of various spatio-temporal extent \citep{diedrich2012cave} and modern terrestrial settings that experience sub-zero conditions can act as a microcosm to show the effect extreme cold can have on scavenging.
Notably, freezing carcasses can become too hard to consume by most vertebrate carnviores \citep{selva2003scavenging}.}

@@

In addition to fending off other vertebrates, scavengers also have to contend with competition from micro-organisms, which requires a specialised physiology. 
Although the findings of \cite{shivik2006vultures} that ``evolutionary pressures favor detection maximizers relative to toxification minimizers in competitive interactions for carcasses." appear sound, the fact remains that overcoming micro-organism toxins is still a beneficial adaptation to any scavenger. 
Avian scavengers have evolved incredibly acidic stomachs that allow them to consume and process putrefied flesh with no ill effects \citep{houston1975digestive,roggenbuck2014microbiome}. 
This adaptation is not restricted to vultures though, \cite{gremillet2012vultures} showed wandering albatrosses (\textit{Diomedea exulans}; so-called ``vultures of the seas'') had an average pH of 1.5, which enables them to consume fisheries discards and squid carcasses. 
There is also evidence of selection for ``toxification minimizers'' beyond birds among the ectotherms.
From our earlier arguments we know that ecthotherms are limited in their ability to find carrion as quickly as endotherms. 
These later arrivers would thus benefit especially from well-developed detoxifying apparatus. 
\cite{shivik2006vultures} suggests that ``specialized oral structures in snakes may have evolved under pressures associated with scavenging."
Moreover, some researchers have charted an evolutionary course from basal fossorial snakes to modern terrestrial species by way of an obligate scavenger intermediate \citep{bauchot2006snakes}. However, snake venom is primary associated with predation in extant species suggesting that if venom played such a role in extinct species it is no longer its main function \citep{casewell2013complex}.
\textcolor{blue}{In water, by contrast, competition with micro-organisms is significantly reduced because of the high pressures and low temperatures \citep{beasley2012carrion}.}
